Time stretching of the GeV emission of GRBs: \\ Fermi-LAT data vs geometrical model

### 2015-12-18 APJ/referee_reply.txt

Dear Editor,
Dear Referee,

We thank you strongly for the comprehensive review of the manuscript which helped us to improve the paper with the account of your comments. We are sorry for the vast delay with the reply due to personal reasons. Please find below answers for the second Referee's comments and comments of the expert on the statistical aspects. We have also studied the literature suggested by the first Referee and referenced when relevant.

Answers to the second Referee report

Q: I think that the consideration of relevant observations and previous papers is weak and needs to be improved. 
A: We agree that many relevant works were indeed missed and now it's fixed in an updated version (see below)

Q: Section 1: "In particular, the total energy emitted in gamma-rays during a burst was found to be similar for the different GRBs within an order of magnitude (Bloom et al. 2003)." More recent papers, with a larger sample and post Fermi, find a wider range of total energy. For example, Frail et al., ApJ, 2012, 746, 156.
"Moreover, the high energy radiation of GRBs is shown to follow the predictions of the synchrotron radiation scenario." This is statement is seriously incomplete and misleading. Many in the field think it likely that the keV-MeV emission is from synchrotron emission, but there there are serious conflicts between at least simple version of synchrotron predictions and observations. There is a vast literature on this topic and numerous papers attempting to resolve the problems. Furthermore, it is unquestionable that multiple emission components are observed in some Fermi GRBs. For example, in addition to the likely-synchrotron component, GRB 090902B has a power-law component (Abdo, A. A. et al. 2009, ApJ, 706, L138).
Yi et al. seems a peculiar reference for spectral lags ... its about short versus long BATSE GRBs. There were numerous papers about BATSE lags during the 1990s. Perhaps find a review article?
"curvature effects. These effects were explored by multiple authors..." There are earlier papers on this topic. e.g., Rees. Fenimore.

A: An introduction is revised according to the comments. References to the recent results are included.

Q: section 2.1 I'm doubtful that the background is constant in time for 24 hours. Perhaps the authors should consult with a member of the LAT team, or justify this assumption.

A: The requirement of the stability of the background is the reason why the "TRANSIENT" event class was not used. For the "SOURCE" class events, the sky background dominates over the misidentification of protons. The corresponding statement is added to section 2.2.

Q: Section 3.
"However, only about 30 of them were detected by the LAT, and only 4 of them were bright in the high energy band. If one extrapolates the uniform jet model to the very high energies, this observation would mean that these groups of bursts are internally different: some of them produce VHE radiation, while the others do not. These differences are hard to explain given that the burst energetics are similar"
There are numerous ways to explain this in addition to jet viewing effects: 1) current evidence is that burst energetics are not similar, 2) some synchrotron spectra might have a cutoff, 3) there might be a range of jet Lorentz factions, leading to a range of cutoffs due to gamma-gamma opacity, 4) the high-energy emission may mostly be due to an additional component (e.g., inverse Compton). The ratio between the nominally synchrotron component and the additional component may vary.
How does this analysis of curvature effects compare to previous published papers?

A: The statement is revised in Section 3. We argue that since the particular reason is non known, one may consider explanation which are based or not based on the internal difference of GRBs.
The difference between our model and previous published papers is now given in the Introduction.

Q: section 4.2
There are many papers on the GRB rate versus z, relating to the star formation rate. Generally there is thought to be a bias towards Pop II and low metallicity stars -- and thus a deficit near z = 0. One of the star formation rate curves could be used.
A: Unfortunately this requires to rerun all the calculation. We will include the star formation data in the next applications of the model.

Q: "since no stars existed at very small redshifts" is "large" meant?
A: corrected

Q: section 5.1
I found this section confusing. It seems to start as a general discussion of how to fit the model to the data, then to implicitly change to fitting the the data of GRB 090926A.
A: Additional explanation is added to Section 5.1. GRB 090926A has the most significant stretching factor and it's explanation is a key to fit all the observations.

Answer to the comments of the expert on the statistical aspects

Q: I see two problems with the use of `KS Probability' in Figures 3-5.

a) The Anderson-Darling statistic is similar to KS but always more sensitive. See https://asaip.psu.edu/Articles/beware-the-kolmogorov-smirnov-test.
b) The KS and AD tests are hypothesis tests, and are not designed for multiple applications for parameter estimation (Stretching Factor). There is no problem with using these statistics as measures of proximity between data and model, but the tabulated significance levels are not applicable (see Web page above). As the original data are time series, simple bootstrapping cannot be performed, so I am not sure how significance levels can be estimated. Some careful thought is needed, or the study should proceed without estimation of the accuracy of the best-fit Stretching Factors.

A: 
We thank the expert for the issues rised. 
a) Specially thanks for interesting reference and suggestion. It's intriguing to test whether AD is stronger in our particular case. Unfortunately, this would require redesign of the code which is optimized for performance with KS. We'll try AD within the next application of our model. 
b) A special bootstrapping procedure should be designed for the precise statistical meaning of the accuracy. Still, the accuracy is given as an estimate, while the main result is the statistical significance of the deviation from stretching factor 1.0. This is estimated by just comparing the data with the expected distribution.

Sincerely yours,
Maxim Piskunov, Grigory Rubtsov

### 2015-12-18 APJ/gammaRaysAPJ.tex
Referee reply draft added. Minor modification to the paper.

Gamma-ray bursts (GRBs) are among the most energetic events in the
Universe. Therefore, the studies of the emission mechanism and
phenomenology of GRBs may provide the new knowledge in particle
physics. Moreover, the GRBs are observed from cosmological distances
and therefore bear an imprint of the late-time evolution of the
Universe. An extensive studies of these explosions led to a number of
interesting results, see~\citet{Vianello:2013ela,Gehrels:2013xd} for a
review. {\bf In particular, there are indications that the total energy emitted in gamma-rays during a
burst is similar for the different GRBs within an order
of magnitude~\citep{postnov1999grb,Bloom:2003wy,Goldstein:2015fib}.
One should keep in mind, though, that there is no agreement on the width of the distribution, and it is in fact model dependent. More recent results, for example, suggest that energetics of different bursts might be different~\citep{racusin2009jet, chandra2012radio}. In this paper, however, we explore the possibility that most of the
bursts have similar energetics in their rest frames to simplify analysis. When more precise energy distribution becomes available, the analysis in this paper may be improved. {\bf Also, it is generally accepted that} the
high energy radiation of GRBs follow{\bf s} the predictions of
the synchrotron radiation scenario~\citep{Wang:2013ptaw}}, {\bf although there are problems with that model~\citep{Daigne:2010fb} and there are bursts in which additional radiation component clearly appears~\citep{Abdo:2009pg} }. Several
observations are related to the temporal variations of spectra. This
way, the spectral lags were found between different low energy bands {\bf \citep{cheng1995time,Wu:1999ms} }and the very-high-energy radiation was discovered to
be extended relative to the x-ray emission
\citep{Castignani:2014gaa,Lange:2013uh,Vianello:2013ela}. In this paper
we go further and compare the temporal extension of GRB gamma-ray
radiation in the two energy bands $\unit[100]{MeV} < E <
\unit[1]{GeV}$ and $E > \unit[1]{GeV}$ (hereafter, low and high-energy
bands) using the data from the Large Area Telescope (LAT) of the Fermi
satellite~\citep{2009ApJ...697.1071A,Ackermann:2012kna}.

@@

We propose that the observed energy dependence of the time profiles
may be explained without introduction of the new effects. We argue
that the stretching may originate from the effects of the jet geometry
called curvature effects. These effects were explored by multiple
authors {\bf \citep{Meszaros:1997je,Soderberg:2001mk,Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba,granot2002off}}. {\bf These
studies either considered the homogeneous
distribution of radiation sources throughout the jet, or assumed dependence of energy and Lorentz factor on the opening angle. Our model is
based on a similar assumption that the highest energy radiators are
concentrated near the axis of the jet and consequently the jet opening
angle depends on energy. The latter is motivated by observations if
all bursts are considered equivalent. In this case the energy
dependence of the jet opening angle resolves the contradiction between
the small fraction of GRBs visible above GeV and the hard
spectrum. We assume the Lorentz factor to be constant throughout the jet for simplicity.}

@@

The exposure map is computed with the \texttt{gtexpcube2} tool. We
compute it with 300 energy bins from \unit[100]{MeV} to
\unit[300]{GeV}. The produced \texttt{expcube} files contain exposures
as functions of energy and location on the sky. We use trilinear
interpolation to compute exposures for all energies and
locations. Knowing the exposures, we use the method introduced in the
appendix of \citet{Rubtsov:2011qq} to estimate the background for low
and high-energy bands. Note, that the ``TRANSIENT'' class photons are
not included in the analysis due to the strong dependence of the
background on the spacecraft position.
{\bf Therefore, the background we estimate is mostly sky background (for the background from misidentification of protons for ``SOURCE'' class events is subdominant), and it remains stable over the period of $24$ hours, making our background estimate valid during the time of the burst.}

@@

First of all, there are around 750 GRBs detected by GBM, half of which
are in the LAT field of view at the moment of observation
\citep{Vianello:2013ela}.  However, only about 30 of them were detected
by the LAT, and only 4 of them were bright in the high energy band.
If one extrapolates the uniform jet model to the very high energies,
this observation would mean that these groups of bursts are internally
different: some of them produce VHE radiation, while the others do
not.  These differences are hard to explain given that the burst
energetics are similar \citep{Bloom:2003wy}.  These differences in
burst counts find natural explanation in our model, in which the
opening angle of a jet is inversely proportional to the energy of the
photons it radiates.  Therefore the most common scenario is that the
off-axis angle of an observer is smaller than the low energy jet
opening angle, but much larger than the opening angle of a high energy
jet.  Due to that, most of the bursts can only be seen at low
energies.  The 4 bursts we study in this paper were seen, according to
our model, from the lowest off-axis angles.

@@

{\it Facilities:} \facility{Fermi}.

>>>>

Gamma-ray bursts (GRBs) are among the most energetic events in the
Universe. Therefore, the studies of the emission mechanism and
phenomenology of GRBs may provide the new knowledge in particle
physics. Moreover, the GRBs are observed from cosmological distances
and therefore bear an imprint of the late-time evolution of the
Universe. An extensive studies of these explosions led to a number of
interesting results, see~\citet{Vianello:2013ela,Gehrels:2013xd} for a
review. {\bf In particular, there are indications that the total
  energy emitted in gamma-rays during a burst is similar for the
  different GRBs within an order of
  magnitude~\citep{postnov1999grb,Bloom:2003wy,Goldstein:2015fib}.
  One should keep in mind, though, that there is no agreement on the
  width of the distribution, and it is in fact model dependent. More
  recent results, for example, suggest that energetics of different
  bursts might be different~\citep{racusin2009jet,
    chandra2012radio}. In this paper, however, we explore the
  possibility that most of the bursts have similar energetics in their
  rest frames to simplify analysis. When more precise energy
  distribution becomes available, the analysis in this paper may be
  improved. {\bf Also, it is generally accepted that} the high energy
  radiation of GRBs follow{\bf s} the predictions of the synchrotron
  radiation scenario~\citep{Wang:2013ptaw}}, {\bf although there are
  problems with that model~\citep{Daigne:2010fb} and there are bursts
  in which additional radiation component clearly
  appears~\citep{Abdo:2009pg}. } Several observations are related to
the temporal variations of spectra. This way, the spectral lags were
found between different low energy bands {\bf
  \citep{cheng1995time,Wu:1999ms} }and the very-high-energy radiation
was discovered to be extended relative to the x-ray emission
\citep{Castignani:2014gaa,Lange:2013uh,Vianello:2013ela}. In this
paper we go further and compare the temporal extension of GRB
gamma-ray radiation in the two energy bands $\unit[100]{MeV} < E <
\unit[1]{GeV}$ and $E > \unit[1]{GeV}$ (hereafter, low and high-energy
bands) using the data from the Large Area Telescope (LAT) of the Fermi
satellite~\citep{2009ApJ...697.1071A,Ackermann:2012kna}.

@@

We propose that the observed energy dependence of the time profiles
may be explained without introduction of the new effects. We argue
that the stretching may originate from the effects of the jet geometry
called curvature effects. These effects were explored by multiple
authors {\bf \citep{Meszaros:1997je,Soderberg:2001mk,Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba,granot2002off}}. {\bf These
studies either considered the homogeneous
distribution of radiation sources throughout the jet, or assumed
dependence of energy and Lorentz factor on the opening angle. The model of this paper is
based on a similar assumption that the highest energy radiators are
concentrated near the axis of the jet and consequently the jet opening
angle depends on energy. The latter is motivated by observations if
all bursts are considered equivalent. In this case the energy
dependence of the jet opening angle resolves the contradiction between
the small fraction of GRBs visible above GeV and the hard
spectrum. We assume the Lorentz factor to be constant throughout the jet for simplicity.}

@@

The exposure map is computed with the \texttt{gtexpcube2} tool. We
compute it with 300 energy bins from \unit[100]{MeV} to
\unit[300]{GeV}. The produced \texttt{expcube} files contain exposures
as functions of energy and location on the sky. We use trilinear
interpolation to compute exposures for all energies and
locations. Knowing the exposures, we use the method introduced in the
appendix of \citet{Rubtsov:2011qq} to estimate the background for low
and high-energy bands. Note, that the ``TRANSIENT'' class photons are
not included in the analysis due to the strong dependence of the
background on the spacecraft position.
{\bf Therefore, the background we estimate is mostly sky background
  for the background from misidentification of protons for ``SOURCE''
  class events is subdominant. Sky background remains stable over the period of $24$ hours, making our estimate valid during the time of the burst.}

@@

First of all, there are around 750 GRBs detected by GBM, half of which
are in the LAT field of view at the moment of observation
\citep{Vianello:2013ela}.  However, only about 30 of them were detected
by the LAT, and only 4 of them were bright in the high energy band.
If one extrapolates the uniform jet model to the very high energies,
this observation would mean that these groups of bursts are internally
different: some of them produce VHE radiation, while the others do
not. {\bf The particular reason for these differences is not
  known. Explanations may include additional spectral components and
  variations of the spectra or jet Lorentz factors within GRB
  population. In this paper we elaborate on explanation, which is not
  based on the difference of GRB. 
}  
The differences in
burst counts find natural explanation in our model, in which the
opening angle of a jet is inversely proportional to the energy of the
photons it radiates.  Therefore the most common scenario is that the
off-axis angle of an observer is smaller than the low energy jet
opening angle, but much larger than the opening angle of a high energy
jet.  Due to that, most of the bursts can only be seen at low
energies.  The 4 bursts we study in this paper were seen, according to
our model, from the lowest off-axis angles.

@@

{\it Facilities:} \facility{Fermi LAT}.
